---
title: "Zillow Comp"
output: html_document
author: "Grace Kim"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
toc: true
knitr:
  opts_chunk:
    echo: FALSE
---



# Statement of Purpose

This report aims to utilize Zillow data to develop predictive models for house price ranges. Targeted at potential homebuyers and real estate stakeholders, my analysis seeks to empower informed decision-making by providing accurate insights into the expected price ranges of houses. The ultimate benefit lies in enabling individuals to navigate the real estate market with confidence and make well-informed investment choices.


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(reticulate)
library(tidyverse)
library(kableExtra)

use_virtualenv("mat434")

```


```{python}

import pandas as pd
from sklearn.model_selection import train_test_split, cross_val_score, GridSearchCV
import matplotlib.pyplot as plt


data = pd.read_csv("C:/Users/grace/OneDrive/Desktop/school/fall 2023/mat434/data.csv.zip")

```


```{r}
py$data %>%
  head() %>%
  kable() %>%
  kable_styling(bootstrap_options = c("hover", "striped"))
```




```{python}

data_rows = data.shape[0]
data_cols = data.shape[1]

```

There are `r py$data_rows` observations and `r py$data_cols` variables in the data.

```{python}

#data.info()

medLotSize = data["lotSizeSqFt"].median()
#data.value_counts("priceRange")
maxLotSize = max(data["lotSizeSqFt"])
minLotSize = min(data["lotSizeSqFt"])

```

The median lot size is `r py$medLotSize` square feet.
The max lot size is `r py$maxLotSize` square feet.
The min lot size is `r py$minLotSize` square feet.


```{python}

train, test = train_test_split(data, train_size = 0.75, random_state = 434)
train.drop(columns='id', inplace=True)
test.drop(columns='id', inplace=True)

```


# Exploratory Data Analysis

I first looked at the distributions of the price ranges and other variables in my data set.

```{python}

sorted_price_ranges = sorted(train['priceRange'].unique())

plt.figure(figsize=(15, 8))
plt.bar(sorted_price_ranges, train['priceRange'].value_counts()[sorted_price_ranges],  color='purple', edgecolor='black')

plt.title('Distribution of Price Range')
plt.xlabel('Price Range')
plt.ylabel('Count')

plt.show()

```

  We can see that there is a pretty even balance between the classes, but the lowest and highest price ranges have less instances. Next, I looked at the distributions of the different predictor variables.
  
  
```{python}

import seaborn as sns

# city
train.value_counts("city")


# home type
train.value_counts("homeType")

# latitude
plt.figure(figsize=(8, 6))
sns.kdeplot(train['latitude'], color='purple', fill=True)

plt.title('Distribution of Latitude')
plt.xlabel('Latitude')
plt.ylabel('Density')

plt.show()

# longitude
plt.figure(figsize=(8, 6))
sns.kdeplot(train['longitude'], color='purple', fill=True)

plt.title('Distribution of Longitude')
plt.xlabel('Longitude')
plt.ylabel('Density')

plt.show()


# garage spaces
plt.figure(figsize=(10, 6))
plt.hist(train['garageSpaces'], bins=20, color='purple', edgecolor='black')

plt.title('Distribution of Garage Spaces')
plt.xlabel('Number of Garage Spaces')
plt.ylabel('Count')

plt.show()

# hasSpa
plt.figure(figsize=(10, 6))

plt.bar(train['hasSpa'].unique(), train['hasSpa'].value_counts(), color='purple', edgecolor='black')

plt.title('Distribution of Spas')
plt.xlabel('Has Spa')
plt.ylabel('Count')

plt.show()

# year built
plt.figure(figsize=(10, 6))

sns.kdeplot(train['yearBuilt'], color='purple', fill=True)

plt.title('Distribution of Year Built')
plt.xlabel('Year Built')
plt.ylabel('Density')

plt.show()


# num of patio and porch features
plt.figure(figsize=(10, 6))

plt.bar(train['numOfPatioAndPorchFeatures'].unique(), train['numOfPatioAndPorchFeatures'].value_counts(), color='purple', edgecolor='black')

plt.title('Distribution of Patio and Porch Features')
plt.xlabel('Number of Patio and Porch Features')
plt.ylabel('Count')

plt.show()

# lot size sq ft ?? NOT WORKING
print("Lot Size plot not displaying properly")
plt.figure(figsize=(10, 6))

plt.hist(train['lotSizeSqFt'], bins=20, color='purple', edgecolor='black')

plt.title('Distribution of Lot Size')
plt.xlabel('Lot Size in Square Feet')
plt.ylabel('Count')

plt.show()

# avg school rating
plt.figure(figsize=(10, 6))

plt.hist(train['avgSchoolRating'], bins=20, color='purple', edgecolor='black')

plt.title('Distribution of School Rating')
plt.xlabel('School Rating')
plt.ylabel('Count')

plt.show()


# median students per teacher
plt.figure(figsize=(10, 6))

plt.hist(train['MedianStudentsPerTeacher'], bins=20, color='purple', edgecolor='black')

plt.title('Distribution of Median Students Per Teacher')
plt.xlabel('Median Students Per Teacher')
plt.ylabel('Count')

plt.show()


# num of bathrooms
plt.figure(figsize=(10, 6))

plt.hist(train['numOfBathrooms'], bins=20, color='purple', edgecolor='black')

plt.title('Distribution of Bathrooms')
plt.xlabel('Number of Bathrooms')
plt.ylabel('Count')

plt.show()

# num of bedrooms
plt.figure(figsize=(10, 6))

plt.hist(train['numOfBedrooms'], bins=20, color='purple', edgecolor='black')

plt.title('Distribution of Bedrooms')
plt.xlabel('Number of Bedrooms')
plt.ylabel('Count')

plt.show()

```

  We see the majority of the homes are in Austin, single family homes, with 0-2 garage spaces. I now will explore how these variables are related to the price of the home.

```{python}

# has spa
price_range_order = sorted_price_ranges

plt.figure(figsize=(12, 8))
sns.countplot(data=train, x='priceRange', hue='hasSpa', palette='pastel', order=price_range_order)

plt.title('Distribution of Spas by Price Range')
plt.xlabel('Price Range')
plt.ylabel('Count')
plt.legend(title='Has Spa', loc='upper right', labels=['No', 'Yes'])

plt.show()



```
We can see the ratio of homes that have a spa increases as the prices increases.

```{python}

plt.figure(figsize=(12, 8))
sns.boxplot(data=train, x='priceRange', y='garageSpaces', palette='pastel', order=price_range_order)

plt.title('Distribution of Garage Spaces by Price Range')
plt.xlabel('Price Range')
plt.ylabel('Garage Spaces')

plt.show()


```
Distribution of garage spaces is pretty even, 0-250000 is slightly lower.


```{python}

plt.figure(figsize=(12, 8))
sns.boxplot(data=train, x='priceRange', y='yearBuilt', palette='pastel', order=price_range_order)

plt.title('Distribution of Year Built by Price Range')
plt.xlabel('Price Range')
plt.ylabel('Year Built')

plt.show()


```
We can see the lower limits of years decreasing as price increases, so the oldest houses are expensive.

```{python}

plt.figure(figsize=(12, 8))
sns.boxplot(data=train, x='priceRange', y='avgSchoolRating', palette='pastel', order=price_range_order)

plt.title('Distribution of School Rating by Price Range')
plt.xlabel('Price Range')
plt.ylabel('Average School Rating')

plt.show()


```

We can see that the average school rating and price range are positively correlated. 

```{python}

plt.figure(figsize=(12, 8))
sns.boxplot(data=train, x='priceRange', y='numOfBathrooms', palette='pastel', order=price_range_order)

plt.title('Distribution of Bathrooms by Price Range')
plt.xlabel('Price Range')
plt.ylabel('Number of Bathrooms')

plt.show()


```

While the lower/mid range houses have an even distribution of bathrooms, the number increases for the higher priced houses.


```{python}

plt.figure(figsize=(12, 8))
sns.boxplot(data=train, x='priceRange', y='MedianStudentsPerTeacher', palette='pastel', order=price_range_order)

plt.title('Distribution of Students Per Teacher by Price Range')
plt.xlabel('Price Range')
plt.ylabel('Median Students Per Teacher')

plt.show()


```

  Like with the number of bathrooms, the median students per teacher also increases for the mid/high range houses.
  

  
